---
title: "Extracting data from the 2022 Uganda Demographic and Health Survey"
author: "Boro, Binya Benson Abe"
format: html
---

1.  Load and install required packages

```{r}
#| label: load and install packages
#| include: false

## Install and load "pacman" to manage required packages shown below

if(!require(pacman)) install.packages("pacman")
p_load(
tidyverse, # data wrangling and piping
janitor, # clean_names() and tabyl()
haven, # read files
here, # obtain files from folders that are not in the working directory
labelled, # work with value labels
arrow # disk based data processing
)
```

2.  Load the data "data/UG_DHS_10PERCENT.dta" which is the 2022 Uganda Demographic and Health Survey (DHS).

```{r}
#| label: load and check raw data
#| include: false

## load and save the data file as "dhs_raw"
dhs_raw <- read_dta(here("data/UG_DHS_10PERCENT.dta"))

## quick view of the dataset
view(dhs_raw)
# Comment: The dataset contains 1436 observations and over 6300 variables.

## create a dataframe for easy understanding of the column variables in the dataset
dhs_df <- generate_dictionary(dhs_raw)

## quick view of the data frame
View(dhs_df)
# Comment: The dataframe provides an easy dictionary for understanding the dataset's attributes
```

3.  Extracting variables of interest

Following Mosley and Chen (1984) framework for analysis of the determinants of child mortality, I grouped the key variables into four domains: Outcome, child-level factors, maternal factors, and household factors.

Outcome: Child mortality, coded as "child is alive" in the dataset

Child level factors: Gender, size, birth order, and others.

Maternal factors: Age, educ, and income

Household factors: Wealth, water, and sanitation

As can be observed in the data framework, (dhs_df), some variables have suffices and others not. I extracted required variables without and with suffices separately before merging them together.

3.1. Extracting variables with no suffices

```{r}
#| label: extract variables with no suffices
#| echo: false

## define key variables, save it as "nsx_dhs"
nsx_dhs <- c(
  "v001", "v002", "v003", # key identifiers
  "v025", # place of residence (1 = Urban, 2= Rural)
  "v190", # wealth index
  "v113", # source of drinking water
  "v012", # respondent's (mother's) age
  "v106" # mother's highes educational level
)

## extract
nsx_extrac <- intersect(nsx_dhs, names(dhs_raw))
no_suffix <- setdiff(nsx_dhs, names(dhs_raw)) # rturning empty cells means that all extracted variables have no suffices
if (length(no_suffix) > 0) {
  message("Non-suffixed vars not found: ",
          paste(no_suffix, collapse = ", "))
}
req_vars <- dhs_raw %>% # code extracts the required variables and save it in "req_vars"
  select(all_of(nsx_extrac))


## check duplicates in extracted variables
req_vars %>%
  count(v001, v002, v003) %>%
  filter(n > 1) %>%
  arrange(desc(n))
# Comment: Duplicate keys exist

## fix duplicate issue and confirm if successful
req_vars_clean <- req_vars %>%
  distinct(v001, v002, v003, .keep_all = TRUE)

## confirm if duplicates still exists
req_vars_clean %>%
  count(v001, v002, v003) %>%
  filter(n > 1) %>%
  nrow()
# comment: Returns value of 0, implies duplicates no more duplicates.
# "req_vars_clean" now contains key variables such as education level, sanitation, location, and wealth index etc.
```

3.2. Extracting variables with several suffices and reshaping them from wide to long

```{r}
#| label: extracting variables with suffices
#| include: false

## define key variables and save them in "sx_dhs"
    sx_dhs <- list(
      "b5_" = "child_alive", # child is alive (0 = no, 1 = yes)
      "b4_" = "child_sex", # sex of child
      "bord_" = "birth_order", # child birth order
      "m14_" = "anc_visits", # number of antenatal visits during pregnancy
      "m19_" = "bweight", # birth weight in kilograms
      "m74_" = "ch_discarge", # child health checked before discharge
      "m4_" = "bflength", # duration of breastfeeding
       "b9_" = "ch_residence" # child lives with whom (0 = with mother, 4 = elsewhere)
    )

## define a function for reshaping to long format
  reshape_sxdhs <- function(dhs_raw, prefix, value_name) {
      dhs_raw %>%
        select(v001, v002, v003, starts_with(prefix)) %>%
        pivot_longer(
          cols         = starts_with(prefix),
          names_to     = "birth_slot",
          names_prefix = prefix,
          values_to    = value_name) %>%
        mutate(birth_slot = as.integer(birth_slot)) %>%
        distinct (v001, v002, v003, birth_slot, .keep_all = TRUE)
    }

## reshape variables from wide to long format
    long_reshape <- imap(sx_dhs, function(value_name, prefix) {

      # Check if any columns with this prefix exist in the dataset
      cols <- names(dhs_raw)[str_detect(names(dhs_raw), paste0("^", prefix))]

      if (length(cols) == 0) {
        message("No columns found for prefix: ", prefix, " — skipping")
        return(NULL)
      }
      message("Reshaping ", length(cols),
              " columns for prefix: ", prefix)
      reshape_sxdhs(dhs_raw, prefix, value_name)
    })

## delete skipped prefixes
    long_reshape <- compact(long_reshape)

## delete main dataset to free up memory space
    rm(dhs_raw) 
    
## define a base variable and combine with other tables of suffixed variables
    comb_sxvars <- long_reshape[["b5_"]] # using "b5_" which indicates child alive as the base 
    for(vars in setdiff(names(long_reshape), "b5_")) {
      comb_sxvars <- comb_sxvars %>% 
        left_join(long_reshape[[vars]],
                  by = c("v001", "v002", "v003", "birth_slot"),
                  relationship = "one-to-one")
    }

## fix duplicates and confirm if successful
    sx_vars_clean <- comb_sxvars %>%
      distinct(v001, v002, v003, birth_slot, .keep_all = TRUE)

## check if duplicates still exist
    sx_vars_clean %>%
      count(v001, v002, v003, birth_slot) %>%
      filter(n > 1) %>%
      nrow() 
    # Comment: Returns 0, means no duplicates

## remove empty cells in the column "child alive" to leave only real births per woman
    sx_vars_clean <- sx_vars_clean %>%
      filter(!is.na(child_alive))
```

3.3. Merging the two data subsets: req_vars_clean - for variables with no suffices and sx_vars_clean - for variables with suffices

```{r}
#| label: merging datasets
#| include: false

## merging variables with and without suffices
    dhs_extrac <- sx_vars_clean %>%
      left_join(req_vars_clean,
                by           = c("v001", "v002", "v003"),
                relationship = "many-to-one") # child births per woman

## verify whether merging is successful
    if (nrow(sx_vars_clean) == nrow(dhs_extrac)) {
      message("Join successful — ", nrow(dhs_extrac), " rows preserved")
    } else {
      message("WARNING — row count changed: ",
              nrow(sx_vars_clean), " before vs ",
              nrow(dhs_extrac), " after")
    }
# Comment: join successful
    
## verify birth slot runs correctly
    dhs_extrac %>%
      summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
      pivot_longer(everything(),
                   names_to  = "variable",
                   values_to = "pct_missing") %>%
      arrange(desc(pct_missing)) %>%
      print(n = Inf)
## view the extracted dataset
    view(dhs_extrac)
## save extracted data as a stata file
    write_dta(dhs_extrac, here("data/dhs_raw_extrac"))
```
