---
title: "data extraction"
author: "Boro, Binya Benson Abe"
format: html
editor: visual
---

1.  Install packages and load the data. The data file "UGKR7BFL_10PERCENT.dta" was downloaded from the Uganda Bureau of Statistics. It consists of 10 percent sample of kid's records in the 2022 Uganda Demography and Health Survey

```{r}
#| label: working directory, packages, and data

## working directory
setwd("/Users/benson/Desktop/econ_analysis/health_analysis")

## install and load packages
if(!require(pacman)) install.packages("pacman") # pacman manages the packages

p_load( # install and load packages
tidyverse,
dplyr,
haven,
here,
labelled,
arrow,
janitor
)

## load the data file: UGKR7BFL_10PERCENT.dta
dhs_KRdt <- read_dta(here("data/UGKR7BFL_10PERCENT.dta"))
view(dhs_KRdt)

## generate a data dictionary to ease search for and understand data attributes
dhs_df <- generate_dictionary(dhs_KRdt)
view(dhs_df)
```

2.  Select relevant variables

I selected the variables of interest following Mosley and Chen's (1984) framework for the analysis child health determinants. They include:

Outcome: child mortality;

Child-level factors: gender, age, size;

Maternal factors: mother's age, education;

Household factors: wealth, water, sanitation

```{r}
#| label: relevant variables
#| echo: false

rel_vars <- dhs_KRdt %>% select(
  # key identifiers
  caseid, # unique mother identifier
  bidx, # child identifier
  v001, # cluster - sampling unit (EA)
  v002, # household within cluster
 
  # outcome variable: child mortality
  b5, # child is alive (1 = yes, 0 = no)
  
  # child-level factors
  b4, # sex
  b8, # age in months
  
  # independent variable
  v106, # mother's level of education
  
  # controls
  v190, # wealth index
  v025, # place of residence (1 = Urban, 2= Rural)
  v024, # region
  v012, # mother's age
  v201, # total children ever born
  
  # others
  v113, # source of drinking water
  m4,  # duration of breastfeeding
  bord, # child birth order
  b11, # preceeding birth interval
  
  # survey design variables
  v005, # sample weight
  v021, # cluster - Primary sampling unit 
  v022 # sample strata
)

# Uniqueness check = TRUE
nrow(dhs_KRdt) == length(unique(paste(dhs_KRdt$caseid, dhs_KRdt$bidx)))
# Note:Unit = child, unique identifier = caseid + bidx

# check whether duplicates exist
rel_vars %>%
  count(caseid,bidx, v001, v002) %>%
  filter(n > 1) %>%
  arrange(desc(n)) # returns "0" rows which implies no duplicates

# save extracted data as a stata file
write_dta(dhs_KRdt, here("data/dhs_extrac.dta"))
```
